WSClean: an implementation of a fast, generic wide-field imager

### 2014-02-17 articles/2014/Article-WSClean/wsclean.tex
Correcting affiliations

\def\ANU{$^{1}$}
\def\CAASTRO{$^{2}$}
\def\ASU{$^{3}$}
\def\CASS{$^{4}$}
\def\Curtin{$^{5}$}
\def\CfA{$^{6}$}
\def\Haystack{$^{7}$}
\def\MIT{$^{8}$}
\def\NRAO{$^{9}$}
\def\RRI{$^{10}$}
\def\SKASA{$^{11}$}
\def\Tata{$^{12}$}
\def\UMelbourne{$^{13}$}
\def\USydney{$^{14}$}
\def\UW{$^{15}$}
\def\UWisc{$^{16}$}
\def\Victoria{$^{17}$}

\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{\hspace{2cm}E-mail:
\url{andre.offringa@anu.edu.au}},
B.~McKinley\ANU$^,$\CAASTRO,
N.~Hurley-Walker\Curtin,
F.~H.~Briggs$^{1,2}$, \newauthor
J.~Rhee\ANU,
L.~Feng,
D.~L.~Kaplan\UWisc,
A.~R.~Neben,
R.~B.~Wayth\Curtin$^,$\CAASTRO, \newauthor
J.~D.~Hughes\ANU,
G.~Bernardi\SKASA,
J.~D.~Bowman\ASU,
F.~Briggs\ANU,
R.~J.~Cappallo\Haystack,
B.~E.~Corey\Haystack, \newauthor
A.~A.~Deshpande\RRI,
D.~Emrich\Curtin,
B.~M.~Gaensler\USydney$^,$\CAASTRO,
R.~Goeke\MIT,
L.~J.~Greenhill\CfA, \newauthor
B.~J.~Hazelton\UW,
J.~N.~Hewitt\MIT,
M.~Johnston-Hollitt\Victoria,
J.~C.~Kasper\CfA, \newauthor
E.~Kratzenberg\Haystack,
C.~J.~Lonsdale\Haystack,
M.~J.~Lynch\Curtin,
S.~R.~McWhirter\Haystack,
D.~A.~Mitchell\CASS$^,$\CAASTRO, \newauthor
M.~F.~Morales\UW,
E.~Morgan\MIT,
D.~Oberoi\Tata,
S.~M.~Ord\Curtin$^,$\CAASTRO,
T.~Prabu\RRI, \newauthor
A.~E.~E.~Rogers\Haystack,
D.~A.~Roshi\NRAO \footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
N.~Udaya~Shankar\RRI,
K.~S.~Srivani\RRI,
R.~Subrahmanyan\RRI$^,$\CAASTRO, \newauthor
S.~J.~Tingay\Curtin$^,$\CAASTRO, 
M.~Waterson\Curtin$^,$\ANU,
R.~L.~Webster\UMelbourne$^,$\CAASTRO,
A.~R.~Whitney\Haystack, \newauthor
A.~Williams\Curtin,
C.~L.~Williams\MIT
\\
$^{1}$Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
$^{2}$ARC Centre of Excellence for All-sky Astrophysics (CAASTRO), Australian National University, Canberra, ACT 2611, Australia \\
$^{3}$School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
$^{4}$CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
$^{5}$International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
$^{6}$Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
$^{7}$MIT Haystack Observatory, Westford, MA 01886, USA\\
$^{8}$Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
$^{9}$National Radio Astronomy Observatory, Charlottesville and Greenbank, USA\\
$^{10}$Raman Research Institute, Bangalore 560080, India\\
$^{11}$Square Kilometre Array South Africa (SKA SA), Cape Town 7405, South Africa\\
$^{12}$National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India\\
$^{13}$School of Physics, The University of Melbourne, Parkville, VIC 3010, Australia\\
$^{14}$Sydney Institute for Astronomy, School of Physics, The University of Sydney, NSW 2006, Australia\\
$^{15}$Department of Physics, University of Washington, Seattle, WA 98195, USA\\
$^{16}$Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
$^{17}$School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
}

@@

\begin{abstract}
We present a new imager implementation that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm. On MWA data, we find it to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution. The performance dependencies of CASA's $w$-projection and WSClean are analysed and analytical functions are derived that describe the required computing cost for both imagers. The difference in computing cost is mainly caused by the efficiency of our new imager on MWA's 27\degree field of view. We predict the computing costs for several other configurations and estimate that our imager is approximately a factor of 2--3 faster on other arrays. Imaging SKA observations is challenging, and SKA1 low will require 3--60 PetaFLOPS of computing power with current techniques. We find that the $w$-snapshot algorithm does not further improve computing requirements. The source code of our new imager is publicly released.
\end{abstract}

@@

\subsection{Optimal snapshot duration}
TODO: need to recalculate this with proper FOV and ZA corrections.
Snapshot imaging, such as the recentring technique described in \S\ref{sec:snapshot-imaging}, can in theory be implemented with either $w$-projection or $w$-stacking. Using the derived formula, we can estimate the cost of making snapshots with both techniques. Snapshot imaging effectively removes the dependency on ZA from $w_{\max}$. Given the snapshot duration $\Delta \tau_\textrm{snapshot}$ and the total integration time $\Delta \tau_\textrm{total}$, the total time cost of inversion becomes
\begin{equation} \label{eq:snapshot-cost}
t_\textrm{snapshot}(\Delta \tau_\textrm{snapshot}) = \frac{\Delta \tau_\textrm{total}}{\Delta \tau_\textrm{snapshot}} t_\textrm{imaging}(w'_{\max},N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
\end{equation}
with
\begin{equation}
w'_{\max} = \frac{1}{\lambda}\max \Delta z \left(1.0 - \cos(\frac{1}{2}(\textrm{FOV} + \omega_E \tau_\textrm{snapshot})\right),
\end{equation}
$\omega_E$ the rotational speed of the Earth and $N'_\textrm{vis} = N_\textrm{vis}\frac{\Delta \tau_\textrm{total}}{\tau_\textrm{snapshot}}$. We have excluded the cost of phase shifting the visibilities and gridding. The cost for regridding with simple nearest neighbour or bilinair interpolation is indeed negligable, although more accurate interpolation (e.g. Lanczos interpolation) can be expensive. Also, our current implementation of the phase-changing program does take a non-negligable time, but this implementation is not optimized and can in theory be implemented in a preprocessing pipeline or on the fly during imaging.

@@

\begin{table*}
\caption{Configurations for which the computational cost are predicted.} \label{tbl:computational-cost-configurations}
\begin{tabular}{l|rrrrrrrrrr}
 Configuration&$\lambda$& FOV  & Beams & Ant & Res & $D$ & $\max \Delta z$ & int. t & BW & $\Delta \nu$ \\
          & (m)       & (FWHM)&       &     &     & (km)                       & (m)             & (s)    & (MHz) &  (kHz) \\
\hline
 GLEAM    &  2   & 24.7\degree& 1    & 128 & 2'   & 2.9 & 5 & 2 & 32 & 40 \\
 EMU      & 0.2  & 1\degree   & 30   &  36 & 10''  & 6 & 0.2 &10 & 300 & 20\\
 MSSS low &  5   & 9.8\degree & 5    &  20 & 100'' & 5 & 2  &10 & 16 & 16 \\
 MSSS high&  2   & 3.8\degree & 5    &  40 & 120'' & 5 & 2  &10 & 16 & 16 \\
 LOFAR LBA NL& 5 & 4.9\degree & 1    &  38 &   3'' &180& 20 & 1 & 96 & 1 \\
 AARTFAAC &  5   & 45\degree  & 1    & 288 &  20'  & 0.3& 0.5 & 1 &  7 & 24 \\
 VLSS     & 4.1  & 14\degree  & 1    &  27 &  80'' &11.1& 0.2 &10 & 1.56 & 12.2 \\
 MeerKAT  & 0.2  &  1\degree  & 1    &  64 &   6'' &  8 & 1   & 0.5 & 750  & 50 \\
% SKA1 AA inner&2 &  5\degree  & 1    & 433 &  15'' &0.6& 0.5 &  10 & 250 & 1 \\
 SKA1 AA core&2  &  5\degree  & 1    & 866 &    3' &  3& 5 & 10 & 250 & 1 \\
 SKA1 AA full&2  &  5\degree  & 1    & 911 &   5'' &100& 50 & 0.6& 250 & 1 \\
\end{tabular}
\end{table*}

@@

An optimisation suggested by \citet{awimager-2013} is to not grid visibilities with $w$-values larger than some value, because this is where most of the computational cost resides when using $w$-projection, while for LOFAR there is little benefit in gridding these samples. In $w$-stacking, the speed gain associated with this optimisation is less significant. Limitting the $w$-values also lowers the snapshots resolution in one direction significantly, because limitting all the long baselines in one direction are no longer gridded, which is often not desirable for the MWA.

\subsection{Estimated computing cost for other telescopes} \label{sec:application-to-nonmwa}
We use the derived functions to estimate the computational cost of the algorithms for configurations of several telescopes. We estimate the cost of imaging an hour of ZA=20\degree data for the following surveys or array configurations: The Galactic and Extragalactic MWA survey (GLEAM); the Evolutionary Map of the Universe (EMU) ASKAP survey \citep{emu-norris-2011}; the low and high bands of the LOFAR Multifrequency Snapshot Sky Survey (MSSS)\footnote{See \url{https://www.astron.nl/radio-observatory/lofar-msss/lofar-msss}}; all Dutch LOFAR stations \citep{lofar-2013}; the Amsterdam—ASTRON Radio Transients Facility and Analysis Centre (AARTFAAC) project\footnote{See \url{http://www.aartfaac.org}}; The VLA Low-Frequency Sky Survey (VLSS, \citealt{vlss-2007}); The Meer Karoo Array Telescope (MeerKAT); and the low-frequency phase~1 aperture arrays of the Square-Kilometre Array, with the full core (3~km) and the core+arms configurations \citep{ska-phase1-2013}. The configurations are summarized in Table~\ref{tbl:computational-cost-configurations}. In multi-beam configurations such as the SKA and EMU configurations, we assume each beam is imaged separately, i.e., the FOV of a single beam is used and the computing cost is multiplied with the number of beams. The image size is determined by 5 times the field of view divided by resolution. The factor of five is to allow better cleaning accuracy and source-position fitting, as is common. The total required computing power is calculated by multiplying the estimated computing time on our test machine with the performance of the machine (138 GFLOPS).

>>>>

\def\ANU{$^{1}$}
\def\CAASTRO{$^{2}$}
\def\Curtin{$^{3}$}
\def\UWisc{$^{4}$}
\def\MIT{$^{5}$}
\def\SKASA{$^{6}$}
\def\ASU{$^{7}$}
\def\Haystack{$^{8}$}
\def\RRI{$^{9}$}
\def\USydney{$^{10}$}
\def\CfA{$^{11}$}
\def\UW{$^{12}$}
\def\Victoria{$^{13}$}
\def\CASS{$^{14}$}
\def\Tata{$^{15}$}
\def\NRAO{$^{16}$}
\def\UMelbourne{$^{17}$}

\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{E-mail:
\url{andre.offringa@anu.edu.au}},
B.~McKinley\ANU$^,$\CAASTRO,
N.~Hurley-Walker\Curtin,
F.~H.~Briggs$^{1,2}$, \newauthor
R.~B.~Wayth\Curtin$^,$\CAASTRO, 
D.~L.~Kaplan\UWisc,
L.~Feng\MIT,
A.~R.~Neben\MIT,
J.~D.~Hughes\ANU, \newauthor
J.~Rhee\ANU,
G.~Bernardi\SKASA,
J.~D.~Bowman\ASU,
R.~J.~Cappallo\Haystack,
B.~E.~Corey\Haystack, \newauthor
A.~A.~Deshpande\RRI, 
D.~Emrich\Curtin,
B.~M.~Gaensler\USydney$^,$\CAASTRO,
R.~Goeke\MIT,
L.~J.~Greenhill\CfA, \newauthor
B.~J.~Hazelton\UW, 
J.~N.~Hewitt\MIT,
M.~Johnston-Hollitt\Victoria,
J.~C.~Kasper\CfA, \newauthor
E.~Kratzenberg\Haystack, 
C.~J.~Lonsdale\Haystack,
M.~J.~Lynch\Curtin,
S.~R.~McWhirter\Haystack,
D.~A.~Mitchell\CASS$^,$\CAASTRO, \newauthor
M.~F.~Morales\UW, 
E.~Morgan\MIT,
D.~Oberoi\Tata,
S.~M.~Ord\Curtin$^,$\CAASTRO,
T.~Prabu\RRI, 
A.~E.~E.~Rogers\Haystack, \newauthor
D.~A.~Roshi\NRAO \footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
N.~Udaya~Shankar\RRI,
K.~S.~Srivani\RRI,
R.~Subrahmanyan\RRI$^,$\CAASTRO, \newauthor
S.~J.~Tingay\Curtin$^,$\CAASTRO, 
M.~Waterson\Curtin$^,$\ANU,
R.~L.~Webster\UMelbourne$^,$\CAASTRO,
A.~R.~Whitney\Haystack, \newauthor
A.~Williams\Curtin,
C.~L.~Williams\MIT
\\
\ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics (CAASTRO), Australian National University, Canberra, ACT 2611, Australia \\
\Curtin{}International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
\UWisc{}Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
\MIT{}Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
\SKASA{}Square Kilometre Array South Africa (SKA SA), Cape Town 7405, South Africa\\
\ASU{}School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
\Haystack{}MIT Haystack Observatory, Westford, MA 01886, USA\\
\RRI{}Raman Research Institute, Bangalore 560080, India\\
\USydney{}Sydney Institute for Astronomy, School of Physics, The University of Sydney, NSW 2006, Australia\\
\CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
\UW{}Department of Physics, University of Washington, Seattle, WA 98195, USA\\
\Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
\CASS{}CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
\Tata{}National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India\\
\NRAO{}National Radio Astronomy Observatory, Charlottesville and Greenbank, USA\\
\UMelbourne{}School of Physics, The University of Melbourne, Parkville, VIC 3010, Australia\\
}

@@

\begin{abstract}
We present a new interferometric imager that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm. On MWA data, we find it to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution. The performance dependencies of CASA's $w$-projection and WSClean are analysed and analytical functions are derived that describe the required computing cost for both imagers. The difference in computing cost is mainly caused by the efficiency of our new imager on MWA's 27\degree field of view. We predict the computing costs for several other configurations and estimate that our imager is approximately a factor of 2--3 faster on other arrays. Imaging SKA observations is challenging, as the full SKA1 low will require 60 PetaFLOPS of computing power with current techniques. We find that the $w$-snapshot algorithm does not further improve computing requirements. The source code of our new imager is publicly released.
\end{abstract}

@@

\subsection{Optimal snapshot duration}
TODO: need to recalculate this with proper FOV and ZA corrections.
Snapshot imaging, such as the recentring technique described in \S\ref{sec:snapshot-imaging}, can be implemented with either $w$-projection or $w$-stacking. Using the derived formula, we can estimate the cost of making snapshots with both techniques. Snapshot imaging effectively removes the dependency on ZA from $w_{\max}$. Given the snapshot duration $\Delta \tau_\textrm{snapshot}$ and the total integration time $\Delta \tau_\textrm{total}$, the total time cost of inversion becomes
\begin{equation} \label{eq:snapshot-cost}
t_\textrm{snapshot}(\Delta \tau_\textrm{snapshot}) = \frac{\Delta \tau_\textrm{total}}{\Delta \tau_\textrm{snapshot}} t_\textrm{imaging}(w'_{\max},N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
\end{equation}
with
\begin{equation}
w'_{\max} = \frac{1}{\lambda}\max \Delta z \left(1.0 - \cos(\frac{1}{2}(\textrm{FOV} + \omega_E \tau_\textrm{snapshot})\right),
\end{equation}
$\omega_E$ the rotational speed of the Earth and $N'_\textrm{vis} = N_\textrm{vis}\frac{\Delta \tau_\textrm{total}}{\tau_\textrm{snapshot}}$. We have excluded the cost of phase shifting the visibilities and gridding. The cost for regridding with simple nearest neighbour or bilinair interpolation is indeed negligable, although more accurate interpolation (e.g. Lanczos interpolation) can be expensive. Also, our current implementation of the phase-changing program does take a non-negligable time, but this implementation is not optimized and can in theory be implemented in a preprocessing pipeline or on the fly during imaging.

@@

\begin{table*}
\caption{Configurations for which the computational cost is predicted.} \label{tbl:computational-cost-configurations}
\begin{tabular}{l|rrrrrrrrrr}
 Configuration&$\lambda$& FOV  & Beams & Ant & Res & $D$ & $\max \Delta z$ & int. t & BW & $\Delta \nu$ \\
          & (m)       & (FWHM)&       &     &     & (km)                       & (m)             & (s)    & (MHz) &  (kHz) \\
\hline
 GLEAM    &  2   & 24.7\degree& 1    & 128 & 2'   & 2.9 & 5 & 2 & 32 & 40 \\
 EMU      & 0.2  & 1\degree   & 30   &  36 & 10''  & 6 & 0.2 &10 & 300 & 20\\
 MSSS low &  5   & 9.8\degree & 5    &  20 & 100'' & 5 & 2  &10 & 16 & 16 \\
 MSSS high&  2   & 3.8\degree & 5    &  40 & 120'' & 5 & 2  &10 & 16 & 16 \\
 LOFAR LBA NL& 5 & 4.9\degree & 1    &  38 &   3'' &180& 20 & 1 & 96 & 1 \\
 AARTFAAC &  5   & 45\degree  & 1    & 288 &  20'  & 0.3& 0.5 & 1 &  7 & 24 \\
 VLSS     & 4.1  & 14\degree  & 1    &  27 &  80'' &11.1& 0.2 &10 & 1.56 & 12.2 \\
 MeerKAT  & 0.2  &  1\degree  & 1    &  64 &   6'' &  8 & 1   & 0.5 & 750  & 50 \\
% SKA1 AA inner&2 &  5\degree  & 1    & 433 &  15'' &0.6& 0.5 &  10 & 250 & 1 \\
 SKA1 AA core&2  &  5\degree  & 1    & 866 &    3' &  3& 5 & 10 & 250 & 1 \\
 SKA1 AA full&2  &  5\degree  & 1    & 911 &   5'' &100& 50 & 0.6& 250 & 1 \\
\end{tabular}
\end{table*}

@@

An optimisation suggested by \citet{awimager-2013} is to not grid visibilities with $w$-values larger than some value, because this is where most of the computational cost resides when using $w$-projection, while for LOFAR there is little benefit in gridding these samples. In $w$-stacking, the speed gain associated with this optimisation is less significant. Limiting the $w$-values also lowers the snapshots resolution in one direction significantly, because the long baselines in one direction are no longer gridded, which is often not desirable for the MWA.

\subsection{Estimated computing cost for other telescopes} \label{sec:application-to-nonmwa}
We use the derived functions to estimate the computational cost of the algorithms for configurations of several telescopes. We estimate the cost of imaging an hour of ZA=20\degree data for the following surveys or array configurations: The Galactic and Extragalactic MWA survey (GLEAM); the Evolutionary Map of the Universe (EMU) ASKAP survey \citep{emu-norris-2011}; the low and high bands of the LOFAR Multifrequency Snapshot Sky Survey (MSSS)\footnote{See \url{https://www.astron.nl/radio-observatory/lofar-msss/lofar-msss}}; all Dutch LOFAR stations \citep{lofar-2013}; the Amsterdam—ASTRON Radio Transients Facility and Analysis Centre (AARTFAAC) project\footnote{See \url{http://www.aartfaac.org}}; The VLA Low-Frequency Sky Survey (VLSS, \citealt{vlss-2007}); The Meer Karoo Array Telescope (MeerKAT); and the low-frequency phase~1 aperture arrays of the Square-Kilometre Array, with the full core (3~km) and the core+arms configurations \citep{ska-phase1-2013}. The configurations are summarized in Table~\ref{tbl:computational-cost-configurations}. In multi-beam configurations such as the SKA and EMU configurations, we assume each beam is imaged separately, i.e., the FOV of a single beam is used and the computing cost is multiplied with the number of beams. The image size is set to 5 times the FOV width divided by the resolution. The factor of five is to allow better cleaning accuracy and source-position fitting, as is common. The total required computing power is calculated by multiplying the estimated computing time on our test machine with the performance of the machine (138 GFLOPS).
